---
title: "Assignment 4"
author: "Roque Arcudia"
date: "August 25, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Take the Boston Housing Data.**

```{r import_data}
library(MASS)
data(Boston)
knitr::kable(head(Boston))
```

**We want to predict median house price. Select the features from the given.**

Calculating the correlation of the features vs median house price

```{r correlation}

corBos <- cor(Boston)
# All cols except mediam value
indBosColNamesIdx <- colnames(corBos)!="medv"
#indBosColNamesIdx

corMedv <- corBos["medv",indBosColNamesIdx]
```

**Arrange the features in the order of their imporance.**

Arranging by absolute value of correlation closer to 1

```{r order_importance}
orderCorMedv <- order(-abs(corMedv))
knitr::kable(corMedv[orderCorMedv])
```

**Divide the data into training and testing.  Fit a linear model.**

Tried to fir using the subset of the highest significance and the subset of highest correlation but the whole set of features gave better RMSE

```{r sample_training}
#subBoston <- head(Boston,100)
subBoston <- Boston
# Highest significance:
#subBoston <- subset(Boston, select=c("lstat","dis","ptratio","rm","nox","medv"))
# Highest correlation: 
#subBoston <- subset(Boston, select=c("lstat","rm","ptratio","indus","tax","medv"))

#knitr::kable(subBoston)
numEntries <- nrow(subBoston)
#numEntries
set.seed(1)
# Selecting 80% of the dataset for training
trainingSamp <- sample(c(1:numEntries),0.8*numEntries,replace=FALSE)
# which not required but easier to visialize
testingSamp <- which(!(c(1:numEntries) %in% trainingSamp))

#trainingSamp
#testingSamp

trainingSet <- subBoston[trainingSamp,]
testingSet <- subBoston[testingSamp,]

#knitr::kable(trainingSet)
#knitr::kable(testingSet)

# medv is a function of everything else
bostonModel <- lm(medv~.,data=trainingSet)

summary(bostonModel)

# Apply the model to the testing set
vals <- predict(bostonModel,testingSet[,attr(bostonModel$terms,"term.labels")])

testingSetPred <- data.frame(testingSet,predicted=vals)
knitr::kable(testingSetPred,row.names=FALSE)

#Quickly get rmse
RMSE <- sqrt(mean((testingSetPred$medv-testingSetPred$predicted)^2))
RMSE

```

**Plot the predicted values of the test data.**

This is multivariable so just plotting the predicted value vs the actual medv

```{r fit_plot}
#plot(bostonModel)

with(testingSetPred, plot(medv,predicted))
abline(a=0,b=1,col="red")

```

**Changed the submission data**
