---
title: "Assignment 4"
author: "Roque Arcudia"
date: "May 28, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r init}
library(dplyr)
inspections <- read.csv("sf-food-csv/inspections_plus.csv",header=TRUE,stringsAsFactors = FALSE)

#products <- read.table("data_products.txt",header=FALSE,sep="#", comment.char = "")

violations <- read.csv("sf-food-csv/violations_plus.csv",header=TRUE,stringsAsFactors = FALSE)
  
#knitr::kable(head(inspections,5),caption = "Inspections")
knitr::kable(head(violations,5),caption = "Violations")

```

## Query 1

1) What is the inspection score distribution like? (inspections_plus.csv)
Expected output - (score, count) - order by score in descending order

```{r q1}
query1 <- na.omit(inspections) %>% group_by(score) %>% summarise(n()) %>% arrange((score))
knitr::kable(head(query1,20),caption = "Violations")
```

## Query 2

2) What is the risk category distribution like? (violations_plus.csv)
Expected output - (risk category, count)

```{r q2}
query2 <- violations %>% group_by(risk_category) %>% summarise(n())
knitr::kable(query2,caption = "Violations")
```

## Query 3

Need clarification

## Query 4

Need clarification

## Query 5

5) Among all the restaurants that got 100 score, what kind of violations did they get (if any)
(inspections_plus.csv, violations_plus.csv)

(Examine "High Risk" violation only)

Expected columns - (business_id, risk_category, date, description)

Note - format the date in (month/day/year)


```{r q5 a}
# Version A any violation that a restaurant that has gotten the 100 score
restaurants100 <- inspections %>% filter(score==100) %>% select(business_id) %>% distinct()
query5a <- violations %>%
  filter(business_id %in% restaurants100$business_id) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5a,caption = "Vio/lations")
```

```{r q5 b}
# Version B the violation has to be the same day it got the 100 score
restaurants100b <- inspections %>% filter(score==100) %>% select(business_id,date) %>% distinct()
query5b <- violations %>%
  semi_join(restaurants100b) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5b,caption = "Query 5b")
```



