---
title: "Assignment 4"
author: "Roque Arcudia"
date: "May 28, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r init}
library(dplyr)
inspections <- read.csv("sf-food-csv/inspections_plus.csv",header=TRUE,stringsAsFactors = FALSE)

#products <- read.table("data_products.txt",header=FALSE,sep="#", comment.char = "")

violations <- read.csv("sf-food-csv/violations_plus.csv",header=TRUE,stringsAsFactors = FALSE)

business <- read.csv("sf-food-csv/businesses_plus.csv",header=TRUE,stringsAsFactors = FALSE)
  
business$postal_code[grepl("CA",business$postal_code)] = NA
business$postal_code[business$postal_code == "941102019"] = "94110"
business$postal_code[business$postal_code == "941033148"] = "94103"

business$postal_code[nchar(business$postal_code) != 5] = NA

business$name[business$business_id==64333]= "The Creme Brulee Cart"
business$name[business$business_id==66922]= "The Creme Brulee Cart"
business$name[business$business_id==67807]= "The Creme Brulee Cart"
business$name[business$business_id==70155]= "The Creme Brulee Cart"
business$name[business$business_id==70705]= "Creme Brulee To Go"
business$name[business$business_id==71259]= "Creme Brulee To Go"

business$name[business$business_id==68091]= "Kien Thanh"

business$name[business$business_id==75448]= "Laguna Cafe"

business$name[business$business_id==81484]= "La Noisette Cafe and Deli"



#knitr::kable(head(inspections,5),caption = "Inspections")
#knitr::kable(head(violations,5),caption = "Violations")
#knitr::kable(head(business,5),caption = "Violations")

#knitr::kable(business[4332,],caption = "Violations")
#knitr::kable(business[,1:7],caption = "Violations")



```

## Query 1

1) What is the inspection score distribution like? (inspections_plus.csv)
Expected output - (score, count) - order by score in descending order

```{r q1}
query1 <- na.omit(inspections) %>% group_by(score) %>% summarise(n()) %>% arrange((score))
knitr::kable(head(query1,20),caption = "Violations")
```

## Query 2

2) What is the risk category distribution like? (violations_plus.csv)
Expected output - (risk category, count)

```{r q2}
query2 <- violations %>% group_by(risk_category) %>% summarise(n())
knitr::kable(query2,caption = "Violations")
```

## Query 3

3) Which 20 businesses got lowest scores?
(inspections_plus.csv, businesses_plus.csv)

(This should be low score rather than lowest score)

Expected columns - (business_id,name,address,city,postal_code,score)

 1       74522           42
 2       68633           46
 3       18480           47
 4         286           49
 5       64154           50
 
```{r q3}

lowest_scores <- head(na.omit(inspections) %>% group_by(business_id) %>% summarise(lowest_score=min(score)) %>% arrange(lowest_score,business_id),20)

business_extract <- business %>% select(business_id,name,address,city,postal_code,latitude,longitude)

query3 <- lowest_scores %>% inner_join(business_extract, by = "business_id")

```

## Query 4

Need clarification

## Query 5

5) Among all the restaurants that got 100 score, what kind of violations did they get (if any)
(inspections_plus.csv, violations_plus.csv)

(Examine "High Risk" violation only)

Expected columns - (business_id, risk_category, date, description)

Note - format the date in (month/day/year)


```{r q5 a}
# Version A any violation that a restaurant that has gotten the 100 score
restaurants100 <- inspections %>% filter(score==100) %>% select(business_id) %>% distinct()
query5a <- violations %>%
  filter(business_id %in% restaurants100$business_id) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5a,caption = "Vio/lations")
```

```{r q5 b}
# Version B the violation has to be the same day it got the 100 score
restaurants100b <- inspections %>% filter(score==100) %>% select(business_id,date) %>% distinct()
query5b <- violations %>%
  semi_join(restaurants100b) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5b,caption = "Query 5b")
```

6) Average inspection score by zip code
(inspections_plus.csv, businesses_plus.csv)

Expected columns - (zip, average score with only two digits after decimal) - order by average inspection score in descending order

```{r q6}

zips <- business %>% select(business_id,postal_code)
zips <- na.omit(zips)

joinScoreZip <- na.omit(inspections) %>% select(business_id, score) %>% left_join(zips,by = "business_id")

joinScoreZip <- na.omit(joinScoreZip)

query6 <- joinScoreZip %>% group_by(postal_code) %>% summarise(avg = round(mean(score),2)) %>% arrange(desc(avg))

knitr::kable(query6,caption = "Query 6")
```

7) Compute the proportion of all businesses in each neighborhood that have incurred at least one of the violations
"High risk vermin infestation"
"Moderate risk vermin infestation"
"Sewage or wastewater contamination”
"Improper food labeling or menu misrepresentation"
"Contaminated or adulterated food”
"Reservice of previously served foods"
"Expected output: zip code, percentage"
This question is asking for each neighborhood, what is the proportion of businesses that have incurred at least one of the above nasty violations

Note: use UDF to determine which violations match with one of the above extremely bad violations

Expected columns - (zip code, total violation count, extreme violation count, proportion with only two digits after decimal)

Order the result by the proportion in descending order

Notes:

This violations ids are NA:

103163
103164
103168

This one is empty description
103140


```{r q7}
badViolations <- c(
  "High risk vermin infestation",
  "Moderate risk vermin infestation",
  "Sewage or wastewater contamination",
  "Improper food labeling or menu misrepresentation",
  "Contaminated or adulterated food",
  "Reservice of previously served foods"
)

#violations %>% filter(business_id==54) %>% mutate(bad_violations = description %in% badViolations) %>% summarise(total=n(),bad=sum(bad_violations),hasBad=any(bad_violations))

#Create a list by business showing how many violations, how many bad violations and if they had any violation

violationSummary <- violations %>% filter(risk_category!="N/A") %>%
  mutate(bad_violations = description %in% badViolations) %>%
  group_by(business_id) %>%
  summarise(violation_count=n(),extreme_violation_count=sum(bad_violations),has_extreme_violation=any(bad_violations))
  

#sum(violationSummary$bad_violations)
#query2 <- violations %>% group_by(description) %>% summarise(n())
#knitr::kable(query2,caption = "Violations")

#knitr::kable(violationSummary,caption = "Violations")

joinBusiness <- zips %>% left_join(violationSummary,by = "business_id")


#clean NA:
joinBusiness$violation_count[is.na(joinBusiness$violation_count)] = 0
joinBusiness$extreme_violation_count[is.na(joinBusiness$extreme_violation_count)] = 0
joinBusiness$has_extreme_violation[is.na(joinBusiness$has_extreme_violation)] = FALSE

query7 <- joinBusiness %>% group_by(postal_code) %>% summarise(
  total_violation_count = sum(violation_count),
  total_extreme_violation_count = sum(extreme_violation_count),
  num = n(),
  proportion_of_business = round(sum(has_extreme_violation)/n(),2)
) %>% arrange(desc(proportion_of_business))


knitr::kable(query7,caption = "Violations")

```



