---
title: "Assignment 4"
author: "Roque Arcudia"
date: "May 28, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r init}
library(dplyr)
inspections <- read.csv("sf-food-csv/inspections_plus.csv",header=TRUE,stringsAsFactors = FALSE)

#products <- read.table("data_products.txt",header=FALSE,sep="#", comment.char = "")

violations <- read.csv("sf-food-csv/violations_plus.csv",header=TRUE,stringsAsFactors = FALSE)

business <- read.csv("sf-food-csv/businesses_plus.csv",header=TRUE,stringsAsFactors = FALSE)
  
business$postal_code[grepl("CA",business$postal_code)] = NA
business$postal_code[business$postal_code == "941102019"] = "94110"
business$postal_code[business$postal_code == "941033148"] = "94103"

business$postal_code[nchar(business$postal_code) != 5] = NA

business$name[business$business_id==64333]= "The Creme Brulee Cart"
business$name[business$business_id==66922]= "The Creme Brulee Cart"
business$name[business$business_id==67807]= "The Creme Brulee Cart"
business$name[business$business_id==70155]= "The Creme Brulee Cart"
business$name[business$business_id==70705]= "Creme Brulee To Go"
business$name[business$business_id==71259]= "Creme Brulee To Go"

business$name[business$business_id==68091]= "Kien Thanh"

business$name[business$business_id==75448]= "Laguna Cafe"

business$name[business$business_id==81484]= "La Noisette Cafe and Deli"



#knitr::kable(head(inspections,5),caption = "Inspections")
#knitr::kable(head(violations,5),caption = "Violations")
#knitr::kable(head(business,5),caption = "Violations")

#knitr::kable(business[4332,],caption = "Violations")
#knitr::kable(business[,1:7],caption = "Violations")



```

## Query 1

1) What is the inspection score distribution like? (inspections_plus.csv)
Expected output - (score, count) - order by score in descending order

```{r q1}
query1 <- na.omit(inspections) %>% group_by(score) %>% summarise(n()) %>% arrange((score))
knitr::kable(head(query1,20),caption = "Violations")
```

## Query 2

2) What is the risk category distribution like? (violations_plus.csv)
Expected output - (risk category, count)

```{r q2}
query2 <- violations %>% group_by(risk_category) %>% summarise(n())
knitr::kable(query2,caption = "Violations")
```

## Query 3

Need clarification

## Query 4

Need clarification

## Query 5

5) Among all the restaurants that got 100 score, what kind of violations did they get (if any)
(inspections_plus.csv, violations_plus.csv)

(Examine "High Risk" violation only)

Expected columns - (business_id, risk_category, date, description)

Note - format the date in (month/day/year)


```{r q5 a}
# Version A any violation that a restaurant that has gotten the 100 score
restaurants100 <- inspections %>% filter(score==100) %>% select(business_id) %>% distinct()
query5a <- violations %>%
  filter(business_id %in% restaurants100$business_id) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5a,caption = "Vio/lations")
```

```{r q5 b}
# Version B the violation has to be the same day it got the 100 score
restaurants100b <- inspections %>% filter(score==100) %>% select(business_id,date) %>% distinct()
query5b <- violations %>%
  semi_join(restaurants100b) %>%
  filter(risk_category=="High Risk") %>%
  select(business_id, risk_category, date, description)
knitr::kable(query5b,caption = "Query 5b")
```

6) Average inspection score by zip code
(inspections_plus.csv, businesses_plus.csv)

Expected columns - (zip, average score with only two digits after decimal) - order by average inspection score in descending order

```{r q6}

zips <- business %>% select(business_id,postal_code)
zips <- na.omit(zips)

joinScoreZip <- na.omit(inspections) %>% select(business_id, score) %>% left_join(zips,by = "business_id")

joinScoreZip <- na.omit(joinScoreZip)

query6 <- joinScoreZip %>% group_by(postal_code) %>% summarise(avg = round(mean(score),2)) %>% arrange(desc(avg))

knitr::kable(query6,caption = "Query 6")
```



